---
title: "LFQ distribution of slim-domain interactions vs background"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: united
author: Bora Uyar
---

`r date()`

# Motivation

The goal of this analysis is to inspect the LFQ scores in the context of SLiM-Domain interactions without considering the SILAC values. 

Approach: From the proteomics screen, find out if LFQ values of
the interactions that can be explained by slim-domain pairs
are higher than the LFQ values of the background distribution.

- For each protein, I find the mean LFQ value of the protein with all peptides (including all genotypes: wt/phos/mut)
- Then I convert the LFQ value to a z-score (proteins with a NA value are converted to minimum LFQ value observed in the screen ~ 15.7). Thus, we can compare LFQ values across proteins. A high z-score means the protein preferentially binds to the target peptide. 
- Finally I check, if the protein-peptide interaction with high z-scores are enriched for slim-domain interactions. 

# Setup

Import required libraries
```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, fig.height = 7, fig.width = 9)
suppressPackageStartupMessages(library(slimR))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(parallel))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(cowplot))

data.table::setDTthreads(12)
ggplot2::theme_set(ggpubr::theme_pubclean())
datadir <- '../data'
figureDir <- '../figures'
tableDir <- '../tables'

knitr::opts_knit$set(root.dir = figureDir)
```

Importing and processing peptide info:
```{r}
peptides <- data.table::fread(file.path(datadir, 'PeptideCandidates1.txt'))
peptides$type <- 'wt'
peptides[grepl('_', peptides$Name)]$type <- 'mut'
peptides[grepl('phos', peptides$Name),]$type <- 'phos'
# remove the leading lower-case 'p' in the phosphorylated form of the sequence
#peptides$Peptide <- gsub("p([STY])", paste0("_phos_", "\\1"), peptides$Peptide)
peptides$Peptide <- gsub("p([STY])", "\\1", peptides$Peptide)
# add leading and trailing "XXX" to the peptide sequence to avoid matching terminal motifs
peptides$Peptide <- paste0('XXX',peptides$Peptide,'XXX')
# assign experiment group ids to make it compatible with Expgroup ids in array data
peptides$Expgroup <- ceiling(peptides$ID/3)*3-2
```

Get ELM classes
```{r elm}
elms <- readRDS(file.path(datadir, 'elm_classes.RDS'))
```

Get uniprot accessions and corresponding gene names from the fasta file of human uniprot sequences
```{r uniprot}
message('Importing fasta sequences')
fasta <- Biostrings::readAAStringSet(file.path(datadir, 'uniprot.9606.Nov_1_2021.fasta'))
uniprot2geneName <- unique(data.table('uniprotAccession' = sub("^sp\\|(.+?)\\|.+$", "\\1", names(fasta)),
                                      'geneName' = sub("^.+ GN=(.+?) .+$", "\\1", names(fasta)),
                                      stringsAsFactors = FALSE))
# simplify fasta headers of fasta data
names(fasta) <- sub("^sp\\|(.+?)\\|.*", "\\1", names(fasta))
```

Get LFQ data, process protein ids, cleanup, assign uniprot ids, get mean LFQ value per interaction from replicates
```{r lfq}
#import LFQ interaction results 
lfqData <- fread(file.path(datadir, 'LFQinteractions.txt'))
# simplify uniprot accession/gene name fields => remove unreviewed ids and keep only one per row
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('lfqData', 'uniprot2geneName'))
s <- pbapply::pbsapply(cl = cl, strsplit(lfqData$Protein.IDs, ";"), function(x) {
  # pick first match
  intersect(x, uniprot2geneName$uniprotAccession)[1]
})
lfqData$uniprotAccession <- s
parallel::stopCluster(cl)

lfqData <- lfqData[!is.na(uniprotAccession)]

lfqData <- melt.data.table(lfqData, id.vars = 'uniprotAccession', 
                measure.vars = grep('wt|mut|phos', colnames(lfqData), value = T))
lfqData$variable <- as.character(lfqData$variable)
lfqData$type <- sub("(wt|phos|mut).+", "\\1", lfqData$variable)
lfqData$Expgroup <- gsub("\\.[0-9]+$", "", gsub("(wt|phos|mut)", "", lfqData$variable))

# get mean LFQ value per interaction 
lfqData <- lfqData[,mean(value, na.rm = T), by = c('uniprotAccession', 'type', 'Expgroup')][!is.na(V1)]
```

Import PFAM annotations from Uniprot and download ELM-PFAM interaction information
```{r elm2pfam}
message("importing PFAM domains")
uniprot2pfam <- readRDS(file.path(datadir, 'uniprot2pfam.RDS'))
uniprot2pfam <- uniprot2pfam[uniprot2pfam$seqnames %in% unique(lfqData$uniprotAccession),]
uniprot2pfam <- unique(subset(uniprot2pfam, select = c('seqnames', 'pfam_acc', 'pfam_name', 'clan')))
colnames(uniprot2pfam) <- c('uniprotAccession', 'PFAM', 'NAME', 'CLAN')
message("importing ELM -> PFAM associations")
# download a table of ELM classes and their cognate PFAM domains
elm2pfam <- readRDS(file.path(datadir, 'elm2pfam.RDS'))

# simplify uniprot2pfam table:
# remove proteins that don't exist in the screen
uniprot2pfam <- uniprot2pfam[uniprotAccession %in% lfqData$uniprotAccession]
```

# Matching SLiMs to PFAM domains

For each peptide, and for each protein in the screen, find which elm-interacting domains exist. 
```{r get_slim_domain_interactions}
# for each peptide, and for each protein in the screen, find which elm-interacting domains exist
slimDomainInteractions <- do.call(rbind, pbapply::pblapply(unique(peptides$Expgroup), function(e) { #for each experiment group
  dt <- do.call(rbind, lapply(c(unique(peptides$type)), function(g) { # for each genotype
    # slims in peptide
    seq <- peptides[Expgroup == e][type == g]$Peptide
    slims <-sort(unique(slimR::searchSLiMs(seq, elms)[['SLiM']]))
    # get list of proteins detected for the experiment group 
    prots <- unique(lfqData[Expgroup == e][type == g]$uniprotAccession)
    
    # find elm-interacting domains among the detected interactors
    dt <- droplevels(merge(uniprot2pfam[uniprotAccession %in% prots],
                           elm2pfam[ELM_identifier %in% slims], by.x = 'NAME', by.y = 'Interaction_Domain_Name'))
    # get a list of slim-domain pairs for each protein
    do.call(rbind, lapply(unique(dt$uniprotAccession), function(x) {
      data.table('uniprotAccession' = x,
                 'slim_domain_pairs' =  paste(paste(dt[uniprotAccession == x]$NAME,
                                                    dt[uniprotAccession == x]$ELM_identifier, sep = "->"),
                                              collapse = '; '),
                 'genotype' = paste0(g, '.slim_domain_pairs'),
                 'Expgroup' = e)
    }))
  }))
}))

slimDomainInteractions <- dcast.data.table(slimDomainInteractions,
                                           Expgroup + uniprotAccession ~ genotype,
                                           value.var = 'slim_domain_pairs')
slimDomainInteractions$Expgroup <- as.character(slimDomainInteractions$Expgroup)
```

# Normalize LFQ score across peptides

Melt the table into matrix form. 
```{r}
x <- dcast(lfqData, Expgroup + type ~ uniprotAccession, value.var = 'V1')
M <- as.matrix(data.frame(x[,-c(1,2)], row.names = paste(x$Expgroup, x$type, sep = '_')))
```

A lot of the proteins have NA values. `quantile(lfqData$V1, 1:1000/1000)` shows that 999/1000 pulldowns return a value higher than 15.7.
So, I set NA LFQ values to this value, basically assuming LFQ of 15 is not pulled down.
```{r}
M[is.na(M)] <- 15.7
```

## Before normalising by peptide

We can observe that some proteins will have high LFQ values for all peptides, while some proteins will have low LFQ values for all peptides. Normalizing the LFQ values to z-scores w.r.t overall interactions, allows us to make them comparable. 

```{r}
ComplexHeatmap::Heatmap(t(M), name = 'LFQ scores across peptides', row_title = 'Proteins', column_title = 'Peptides', use_raster = F, show_row_names = F, show_column_names = F)
```

## After normalising by peptide
Scaling each row, assigns z-scores to each protein (normalizing w.r.t other peptides)
```{r }
M <- t(scale(M)) # (each protein is normalised w.r.t other peptides)
ComplexHeatmap::Heatmap(M, name = 'LFQ scores across peptides', row_title = 'Proteins', column_title = 'Peptides', use_raster = F, show_row_names = F, show_column_names = F)
write.table(M, file = file.path(tableDir, 
                                'LFQinteractions.scaled_by_peptide.tsv'),
            sep = '\t', quote = F)
```

Now, we can see blocks of red cells that have high z-scores for groups of proteins. 
Now, we check if known SLiM-Domain interactions are enriched in such "red" blocks. 

## Combining LFQ data with slim-domain pairs

I obtain the scaled/normalized values of all pairs of peptides/proteins and check the 
distribution of z-scores for those categorized by whether the slim-domain interactions 
have higher z-scores overall. 

```{r}
lfqScaled <- data.table(melt(M))
lfqScaled$Expgroup <- unlist(lapply(strsplit(as.character(lfqScaled$Var2), "_"), function(a) a[1]))
lfqScaled$type <- unlist(lapply(strsplit(as.character(lfqScaled$Var2), "_"), function(a) a[2]))
lfqScaled$Var2 <- NULL
colnames(lfqScaled)[1] <- 'uniprotAccession'

# see if interactions explained by slim-domain pairs
# have higher LFQ values
lfqMerged <- merge(lfqScaled, slimDomainInteractions,
           by = c('uniprotAccession', 'Expgroup'), all = T)
lfqMerged$paired = FALSE
lfqMerged[type == 'wt'][!is.na(wt.slim_domain_pairs)]$paired <- TRUE
lfqMerged[type == 'mut'][!is.na(mut.slim_domain_pairs)]$paired <- TRUE
lfqMerged[type == 'phos'][!is.na(phos.slim_domain_pairs)]$paired <- TRUE

write.table(lfqMerged, file = file.path(tableDir, 
                                        'LFQscaled_slim_domain_interactions.tsv'), sep = '\t', quote = F)

```

We can see the LFQ values for paired interactions are higher than the rest. However, this seems to be true for all genotypes (phos/wt/mut). This can be explained by the fact that slim-domain pairs we defined may not be phosphorylation dependent docking motifs. For example, a lot of slim-domain pairs are defined where the peptide has a "MOD" type motif (e.g. MOD_GSK3) and the interacting protein has a Pkinase domain. This is a valid ELM-PFAM pair, however, this doesn't mean the protein with the kinase domain will dock onto this peptide.

```{r}
ggplot(lfqMerged, aes(x = value)) + geom_density(aes(fill = paired), alpha = 0.4) + 
  facet_wrap(~ type, ncol = 1) + labs(x = 'LFQ z-score', fill = 'SLiM-Domain Interaction')
ggsave(filename = file.path(figureDir, 
                            'lfq_zscore_vs_slimdomain_interactions.pdf'), width = 10, height = 8)
```

Therefore, we categorize slim-domain pairs by if the peptide has a phosphorylation dependent docking/ligand/degron motif and the interaction partner has a cognate domain. If we focus on slim-domain pairs, where the peptide has a phosphorylation dependent docking/ligand/degron motif (excluding modification/targeting/cleavage motifs), then we see the normalized LFQ scores for phos peptides are higher than both wt/mut peptides. 

```{r}
pat <- "(\\(\\[[ST]+\\]\\)|\\([YTS]\\))"
e <- names(grep(pat, unlist(elms[grep('MOD|TRG|CLV', names(elms), invert = T)]), value = T))
lfqMerged$phosDependentMotifs <- unlist(lapply(strsplit(lfqMerged$phos.slim_domain_pairs, "; "), function(x) sum(gsub("^.+->", "", x) %in% e)))

ggboxplot(lfqMerged[phosDependentMotifs > 0], x = 'type', y = 'value', color = 'type',
         add = 'jitter') + stat_compare_means(method.args = list('alternative' = 'greater'), 
                                               comparisons = list(c('phos', 'wt'),
                                                                  c('phos', 'mut'))) + 
  labs(title = 'LFQ score distributions of interactions explained by\nphospho-dependent-SLiM-Domain pairs',
       x = 'peptide type', y = 'LFQ score')
ggsave(filename = file.path(figureDir, 
                            'lfq_zscore_phos_vs_wt_vs_mut.doc_lig_deg.pdf'), width = 5, height = 7)

```

If we also exclude the degron motifs, which require double phosphorylation, then we see an even stronger difference. 
```{r}
ggboxplot(lfqMerged[phosDependentMotifs > 0][grep('DEG', phos.slim_domain_pairs, invert = T)], x = 'type', y = 'value', color = 'type',
         add = 'jitter') + stat_compare_means(method.args = list('alternative' = 'greater'), 
                                               comparisons = list(c('phos', 'wt'),
                                                                  c('phos', 'mut'))) +
    labs(title = 'LFQ score distributions of interactions explained by\nphospho-dependent-SLiM-Domain pairs',
       x = 'peptide type', y = 'LFQ score')
ggsave(filename = file.path(figureDir, 
                            'lfq_zscore_phos_vs_wt_vs_mut.doc_lig.pdf'), width = 5, height = 7)
```

# Summary

When I look into the LFQ values normalized across peptides, I see a higher z-score
for slim-domain paired protein-peptide interactions. However, this is true for 
all genotypes. I would expect to see that e.g. phosphopeptides have a higher z-score
compared to wild-type or mutant peptides. Following up on this, we decided to 
categorize such slim-domain pairs by if the interaction is expected to be a phosphorylation-dependent docking interaction (excluding targeting/modification motifs). Then we observe that the 
phos peptides have higher scores than both wt/mut peptides. An interesting group are the degrons, which usually require double phosphorylation sites. The difference between phos and wt/mut get 
even bigger if we exclude degrons. This could be because degrons require double phosphorylation sites, while the peptides in the screen were designed with only a single phosphorylation. 

# Session Info
```{r}
print(sessionInfo())
```

